---
title: "Introduction to lwpls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to lwpls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
pkgs <- c("modeldata", "rsample", "tune", "workflows", "yardstick", "ggplot2")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = rlang::is_installed(pkgs)
)
```

Locally-weighted partial least squares (LW-PLS) is a *just-in-time* learning
method. Instead of fitting a single global PLS model, LW-PLS builds a new
weighted PLS model around every sample to be predicted. Training samples that
are similar to the query get large weights and dissimilar ones get small
weights. The result is a locally linear model that can follow nonlinear and
drifting relationships, which is why LW-PLS is widely used for soft sensors and
for spectroscopic calibration.

The lwpls package implements LW-PLS natively (the prediction kernel is written
in C++) and integrates it with tidymodels:

* `lwpls()` is a [parsnip](https://parsnip.tidymodels.org) model specification
  for regression and classification;
* `localization()` is a [dials](https://dials.tidymodels.org) tuning parameter;
* [tune](https://tune.tidymodels.org) evaluates all values of `num_comp` from a
  single model fit.

```{r}
#| label: setup
#| message: false
library(lwpls)
library(parsnip)
library(rsample)
library(tune)
library(workflows)
library(yardstick)
library(ggplot2)
```

## The algorithm

For a query sample $x_q$, the similarity weight of the training sample $x_i$ is

$$
\omega_i = \exp\left(-\frac{d_i}{\sigma_d \, \varphi}\right), \qquad
d_i = \lVert x_i - x_q \rVert,
$$

where $\sigma_d$ is the standard deviation of the distances $d_i$ and
$\varphi > 0$ is the **localization** parameter. The training data are centered
with the weighted means, a PLS model with `num_comp` components is fitted using
the weights $\omega_i$, and the query is projected onto that local model to get
its prediction.

The two main hyperparameters are therefore:

* `num_comp`: the number of PLS components of the local models;
* `localization`: small values make the models very local (only the nearest
  samples matter); large values make all weights similar, so LW-PLS tends to a
  global PLS model.

Optionally, `neighbors` restricts each local model to the nearest training
samples, which reduces the cost of each local model for large training sets.

## Regression: fat-free water content of meat from NIR spectra

The `meats` data from the modeldata package contain 100 near-infrared
absorbance values for 215 meat samples. We predict the water content.

```{r}
#| label: data
data(meats, package = "modeldata")
meats <- meats[, !names(meats) %in% c("fat", "protein")]

set.seed(123)
meats_split <- initial_split(meats, prop = 0.75)
meats_train <- training(meats_split)
meats_test <- testing(meats_split)
meats_folds <- vfold_cv(meats_train, v = 5)
```

The model specification marks both hyperparameters for tuning. Since LW-PLS
calculates its local models at prediction time, all values of `num_comp` are
obtained from the same model, and tune fits only one model per value of
`localization` (the "submodel trick").

```{r}
#| label: tune
lwpls_spec <-
  lwpls(num_comp = tune(), localization = tune()) |>
  set_mode("regression")

lwpls_wflow <- workflow(water ~ ., lwpls_spec)

lwpls_grid <- dials::grid_regular(
  dials::num_comp(c(1, 15)),
  localization(),
  levels = c(num_comp = 15, localization = 8)
)

lwpls_res <- tune_grid(
  lwpls_wflow,
  resamples = meats_folds,
  grid = lwpls_grid,
  metrics = metric_set(rmse, rsq)
)

show_best(lwpls_res, metric = "rmse", n = 5)
```

```{r}
#| label: tune-plot
#| fig-alt: "Cross-validated RMSE versus the number of components, one line per localization value."
autoplot(lwpls_res, metric = "rmse") +
  scale_color_viridis_d(option = "mako", end = 0.9) +
  theme_bw()
```

Very local models (small `localization`) quickly over-fit, while very large
values reproduce global PLS. The best models lie in between.

We finalize the workflow with the best parameters and evaluate it on the test
set:

```{r}
#| label: last-fit
best_params <- select_best(lwpls_res, metric = "rmse")
best_params

lwpls_final <-
  lwpls_wflow |>
  finalize_workflow(best_params) |>
  last_fit(meats_split)

collect_metrics(lwpls_final)
```

### Comparison with global PLS

With a very large localization, all training samples get the same weight and
LW-PLS reduces to ordinary PLS. This gives a convenient baseline:

```{r}
#| label: global-pls
pls_wflow <- workflow(
  water ~ .,
  lwpls(num_comp = tune(), localization = 1e6) |> set_mode("regression")
)
pls_res <- tune_grid(
  pls_wflow,
  resamples = meats_folds,
  grid = data.frame(num_comp = 1:15),
  metrics = metric_set(rmse, rsq)
)

pls_final <-
  pls_wflow |>
  finalize_workflow(select_best(pls_res, metric = "rmse")) |>
  last_fit(meats_split)

collect_metrics(pls_final)
```

The local models clearly improve on the global calibration.

```{r}
#| label: obs-pred
#| fig-alt: "Observed versus predicted water content on the test set for global PLS and LW-PLS."
test_preds <- rbind(
  data.frame(model = "Global PLS", collect_predictions(pls_final)),
  data.frame(model = "LW-PLS", collect_predictions(lwpls_final))
)

ggplot(test_preds, aes(water, .pred)) +
  geom_abline(color = "grey50", linetype = 2) +
  geom_point(alpha = 0.7) +
  facet_wrap(~model) +
  coord_obs_pred() +
  labs(x = "Observed water (%)", y = "Predicted water (%)") +
  theme_bw()
```

## Classification

For a factor outcome, LW-PLS models the class indicators (LW-PLS-DA). The
predicted class is the one with the largest predicted indicator, and the
indicators, truncated to $[0, 1]$, give the class probabilities.

```{r}
#| label: classification
set.seed(1)
iris_split <- initial_split(iris, strata = Species)

iris_fit <-
  lwpls(num_comp = 2, localization = 0.5) |>
  set_mode("classification") |>
  fit(Species ~ ., data = training(iris_split))

iris_preds <- augment(iris_fit, testing(iris_split))
head(iris_preds[, 1:5])
accuracy(iris_preds, truth = Species, estimate = .pred_class)
```

## Using the engine directly

The fitting function `lwpls_fit()` can be used without parsnip. It has matrix,
data frame, formula and recipe interfaces, and its `predict()` method can use any
number of components:

```{r}
#| label: engine
fit <- lwpls_fit(water ~ ., data = meats_train, num_comp = 10, localization = 0.5)
fit

predict(fit, head(meats_test))
predict(fit, head(meats_test), num_comp = 3)
```

Several numbers of components can be predicted at once from a parsnip model
with `multi_predict()`:

```{r}
#| label: multi-predict
parsnip_fit <- extract_fit_parsnip(lwpls_final)
multi_predict(parsnip_fit, head(meats_test), num_comp = c(5, 10, 15))$.pred[[1]]
```

## Engine arguments

The `"lwpls"` engine standardizes the predictors (and numeric outcomes) before
computing distances by default. For data measured on a common scale, such as
spectra, this can be switched off:

```{r}
#| label: engine-args
lwpls(num_comp = 10, localization = 0.5) |>
  set_mode("regression") |>
  set_engine("lwpls", scale = FALSE) |>
  translate()
```

## References

* Kim, S., Kano, M., Nakagawa, H. and Hasebe, S. (2011). Estimation of active
  pharmaceutical ingredients content using locally weighted partial least
  squares and statistical wavelength selection. *International Journal of
  Pharmaceutics*, 421(2), 269--274.
* Kaneko, H. Locally-weighted partial least squares (LWPLS).
  <https://datachemeng.com/locallyweightedpartialleastsquares/>
* Dayal, B. S. and MacGregor, J. F. (1997). Improved PLS algorithms.
  *Journal of Chemometrics*, 11(1), 73--85.
