---
title: "Introduction to lwpls"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to lwpls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
pkgs <- c("modeldata", "rsample", "tune", "workflows", "yardstick", "ggplot2")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = rlang::is_installed(pkgs)
)
```

Locally-weighted partial least squares (LW-PLS) is a *just-in-time* learning
method. Instead of fitting a single global PLS model, LW-PLS builds a new
weighted PLS model around every sample to be predicted. Training samples that
are similar to the query get large weights and dissimilar ones get small
weights. The result is a locally linear model that can follow nonlinear and
drifting relationships, which is why LW-PLS is widely used for soft sensors and
for spectroscopic calibration.

The lwpls package implements LW-PLS natively (the prediction kernel is written
in C++) and integrates it with tidymodels:

* `lwpls()` is a [parsnip](https://parsnip.tidymodels.org) model specification
  for regression and classification;
* `localization()` is a [dials](https://dials.tidymodels.org) tuning parameter;
* [tune](https://tune.tidymodels.org) evaluates all values of `num_comp` from a
  single model fit.

```{r}
#| label: setup
#| message: false
library(lwpls)
library(parsnip)
library(rsample)
library(tune)
library(workflows)
library(yardstick)
library(ggplot2)
```

## The algorithm

For a query sample $x_q$, the similarity weight of the training sample $x_i$ is

$$
\omega_i = \exp\left(-\frac{d_i}{\sigma_d \, \varphi}\right), \qquad
d_i = \lVert x_i - x_q \rVert,
$$

where $\sigma_d$ is the standard deviation of the distances $d_i$ and
$\varphi > 0$ is the **localization** parameter. The training data are centered
with the weighted means, a PLS model with `num_comp` components is fitted using
the weights $\omega_i$, and the query is projected onto that local model to get
its prediction.

The two main hyperparameters are therefore:

* `num_comp`: the number of PLS components of the local models;
* `localization`: small values make the models very local (only the nearest
  samples matter); large values make all weights similar, so LW-PLS tends to a
  global PLS model.

Optionally, `neighbors` restricts each local model to the nearest training
samples, which reduces the cost of each local model for large training sets.

## Regression: fat-free water content of meat from NIR spectra

The `meats` data from the modeldata package contain 100 near-infrared
absorbance values for 215 meat samples. We predict the water content.

```{r}
#| label: data
data(meats, package = "modeldata")
meats <- meats[, !names(meats) %in% c("fat", "protein")]

set.seed(123)
meats_split <- initial_split(meats, prop = 0.75)
meats_train <- training(meats_split)
meats_test <- testing(meats_split)
meats_folds <- vfold_cv(meats_train, v = 5)
```

The model specification marks both hyperparameters for tuning. Since LW-PLS
calculates its local models at prediction time, all values of `num_comp` are
obtained from the same model, and tune fits only one model per value of
`localization` (the "submodel trick").

```{r}
#| label: tune
lwpls_spec <-
  lwpls(num_comp = tune(), localization = tune()) |>
  set_mode("regression")

lwpls_wflow <- workflow(water ~ ., lwpls_spec)

lwpls_grid <- dials::grid_regular(
  dials::num_comp(c(1, 15)),
  localization(),
  levels = c(num_comp = 15, localization = 8)
)

lwpls_res <- tune_grid(
  lwpls_wflow,
  resamples = meats_folds,
  grid = lwpls_grid,
  metrics = metric_set(rmse, rsq)
)

show_best(lwpls_res, metric = "rmse", n = 5)
```

```{r}
#| label: tune-plot
#| fig-alt: "Cross-validated RMSE versus the number of components, one line per localization value."
autoplot(lwpls_res, metric = "rmse") +
  scale_color_viridis_d(option = "mako", end = 0.9) +
  theme_bw()
```

Very local models (small `localization`) quickly over-fit, while very large
values reproduce global PLS. The best models lie in between.

We finalize the workflow with the best parameters and evaluate it on the test
set:

```{r}
#| label: last-fit
best_params <- select_best(lwpls_res, metric = "rmse")
best_params

lwpls_final <-
  lwpls_wflow |>
  finalize_workflow(best_params) |>
  last_fit(meats_split)

collect_metrics(lwpls_final)
```

### Comparison with global PLS

With a very large localization, all training samples get the same weight and
LW-PLS reduces to ordinary PLS. This gives a convenient baseline:

```{r}
#| label: global-pls
pls_wflow <- workflow(
  water ~ .,
  lwpls(num_comp = tune(), localization = 1e6) |> set_mode("regression")
)
pls_res <- tune_grid(
  pls_wflow,
  resamples = meats_folds,
  grid = data.frame(num_comp = 1:15),
  metrics = metric_set(rmse, rsq)
)

pls_final <-
  pls_wflow |>
  finalize_workflow(select_best(pls_res, metric = "rmse")) |>
  last_fit(meats_split)

collect_metrics(pls_final)
```

The local models clearly improve on the global calibration.

```{r}
#| label: obs-pred
#| fig-alt: "Observed versus predicted water content on the test set for global PLS and LW-PLS."
test_preds <- rbind(
  data.frame(model = "Global PLS", collect_predictions(pls_final)),
  data.frame(model = "LW-PLS", collect_predictions(lwpls_final))
)

ggplot(test_preds, aes(water, .pred)) +
  geom_abline(color = "grey50", linetype = 2) +
  geom_point(alpha = 0.7) +
  facet_wrap(~model) +
  coord_obs_pred() +
  labs(x = "Observed water (%)", y = "Predicted water (%)") +
  theme_bw()
```

## Classification

For a factor outcome, LW-PLS models the class indicators (LW-PLS-DA). The
predicted class is the one with the largest predicted indicator, and the
indicators, truncated to $[0, 1]$, give the class probabilities.

Because every local model is linear but fitted around its own query, LW-PLS-DA
can follow curved class boundaries that a global PLS-DA model cannot. The
`parabolic` data from modeldata illustrate this: two classes separated by a
parabola in the plane of the predictors `X1` and `X2`.

```{r}
#| label: parabolic-data
data(parabolic, package = "modeldata")

set.seed(91)
para_split <- initial_split(parabolic, strata = class)
para_train <- training(para_split)
para_test <- testing(para_split)
para_folds <- vfold_cv(para_train, v = 5, strata = class)
```

With two predictors, local models have at most two components, so we tune
`localization` only:

```{r}
#| label: classification-tune
lwplsda_wflow <- workflow(
  class ~ .,
  lwpls(num_comp = 2, localization = tune()) |> set_mode("classification")
)

lwplsda_res <- tune_grid(
  lwplsda_wflow,
  resamples = para_folds,
  grid = data.frame(localization = 2^seq(-6, 5)),
  metrics = metric_set(roc_auc, accuracy)
)

show_best(lwplsda_res, metric = "roc_auc", n = 3)
```

```{r}
#| label: classification-tune-plot
#| fig-height: 3.5
#| fig-alt: "Cross-validated ROC AUC and accuracy versus the localization parameter on a log-2 scale. Both decrease as localization increases."
autoplot(lwplsda_res) +
  theme_bw()
```

Performance drops steadily as the models become more global. We compare the
tuned LW-PLS-DA model with global PLS-DA (a very large `localization`) on the
test set:

```{r}
#| label: classification-fit
lwplsda_final <-
  lwplsda_wflow |>
  finalize_workflow(select_best(lwplsda_res, metric = "roc_auc")) |>
  last_fit(para_split, metrics = metric_set(roc_auc, accuracy))

plsda_final <-
  workflow(
    class ~ .,
    lwpls(num_comp = 2, localization = 1e6) |> set_mode("classification")
  ) |>
  last_fit(para_split, metrics = metric_set(roc_auc, accuracy))

rbind(
  data.frame(model = "LW-PLS-DA", collect_metrics(lwplsda_final)),
  data.frame(model = "Global PLS-DA", collect_metrics(plsda_final))
)[, c("model", ".metric", ".estimate")]
```

The decision regions show why. The global model can only separate the classes
with a straight line, while the local models bend the boundary along the
parabola. The shading is the predicted probability of `Class1`, the black line
is the 0.5 contour, and the points are the test samples:

```{r}
#| label: classification-regions
#| fig-height: 4
#| fig-alt: "Predicted probability of Class1 over the predictor plane for global PLS-DA and LW-PLS-DA, with test samples. The global boundary is a straight line; the LW-PLS-DA boundary follows the parabola."
para_grid <- expand.grid(
  X1 = seq(min(parabolic$X1), max(parabolic$X1), length.out = 100),
  X2 = seq(min(parabolic$X2), max(parabolic$X2), length.out = 100)
)

region_probs <- function(final, model) {
  probs <- predict(extract_workflow(final), para_grid, type = "prob")
  data.frame(para_grid, model = model, .pred_Class1 = probs$.pred_Class1)
}

regions <- rbind(
  region_probs(plsda_final, "Global PLS-DA"),
  region_probs(lwplsda_final, "LW-PLS-DA")
)

ggplot(regions, aes(X1, X2)) +
  geom_raster(aes(fill = .pred_Class1), alpha = 0.8) +
  geom_contour(aes(z = .pred_Class1), breaks = 0.5, color = "black") +
  geom_point(data = para_test, aes(shape = class), color = "white", size = 2.6) +
  geom_point(data = para_test, aes(shape = class, color = class), size = 1.6) +
  facet_wrap(~model) +
  scale_fill_distiller(palette = "RdBu", direction = 1, limits = c(0, 1)) +
  scale_color_manual(values = c(Class1 = "#08306B", Class2 = "#67000D")) +
  coord_equal(expand = FALSE) +
  labs(fill = "Probability\nof Class1", color = NULL, shape = NULL) +
  theme_bw()
```

The ROC curves on the test set summarize the difference across all probability
thresholds:

```{r}
#| label: classification-roc
#| fig-height: 4
#| fig-alt: "Test-set ROC curves of LW-PLS-DA and global PLS-DA. The LW-PLS-DA curve is closer to the top-left corner."
test_roc <- function(final, model) {
  roc <- roc_curve(collect_predictions(final), truth = class, .pred_Class1)
  data.frame(model = model, roc)
}

roc_data <- rbind(
  test_roc(lwplsda_final, "LW-PLS-DA"),
  test_roc(plsda_final, "Global PLS-DA")
)

ggplot(roc_data, aes(1 - specificity, sensitivity, color = model)) +
  geom_abline(color = "grey50", linetype = 2) +
  geom_path(linewidth = 0.8) +
  scale_color_manual(values = c("LW-PLS-DA" = "#08519C", "Global PLS-DA" = "#969696")) +
  coord_equal() +
  labs(color = NULL) +
  theme_bw()
```

Finally, the confusion matrix of the LW-PLS-DA model on the test set:

```{r}
#| label: classification-confusion
#| fig-width: 4
#| fig-height: 3.5
#| fig-alt: "Confusion matrix heatmap of LW-PLS-DA on the test set."
collect_predictions(lwplsda_final) |>
  conf_mat(truth = class, estimate = .pred_class) |>
  autoplot(type = "heatmap")
```

Note that in regions without training data, such as the upper-left corner, the
local models extrapolate from the nearest samples, so their predictions should
not be over-interpreted.

For more than two classes, the model works the same way and returns one
`.pred_{level}` probability column per class.

## Using the engine directly

The fitting function `lwpls_fit()` can be used without parsnip. It has matrix,
data frame, formula and recipe interfaces, and its `predict()` method can use any
number of components:

```{r}
#| label: engine
fit <- lwpls_fit(water ~ ., data = meats_train, num_comp = 10, localization = 0.5)
fit

predict(fit, head(meats_test))
predict(fit, head(meats_test), num_comp = 3)
```

Several numbers of components can be predicted at once from a parsnip model
with `multi_predict()`:

```{r}
#| label: multi-predict
parsnip_fit <- extract_fit_parsnip(lwpls_final)
multi_predict(parsnip_fit, head(meats_test), num_comp = c(5, 10, 15))$.pred[[1]]
```

## Engine arguments

The `"lwpls"` engine standardizes the predictors (and numeric outcomes) before
computing distances by default. For data measured on a common scale, such as
spectra, this can be switched off:

```{r}
#| label: engine-args
lwpls(num_comp = 10, localization = 0.5) |>
  set_mode("regression") |>
  set_engine("lwpls", scale = FALSE) |>
  translate()
```

## References

* Kim, S., Kano, M., Nakagawa, H. and Hasebe, S. (2011). Estimation of active
  pharmaceutical ingredients content using locally weighted partial least
  squares and statistical wavelength selection. *International Journal of
  Pharmaceutics*, 421(2), 269--274.
* Dayal, B. S. and MacGregor, J. F. (1997). Improved PLS algorithms.
  *Journal of Chemometrics*, 11(1), 73--85.
