---
title: "Choosing a similarity index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a similarity index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.2,
  fig.retina = 2,
  fig.align = "center",
  out.width = "100%"
)
```

LW-PLS fits a separate model around every query sample, and that model is
only as good as the training samples it is built on. The `similarity`
argument decides which samples these are: it sets how the distance between a
training sample and the query is measured before the distance is turned into
a weight. This article explains the two indexes available in lwpls, shows
with simulated data when each one works and when it fails, and ends with
practical advice for choosing between them.

```{r}
#| label: setup
#| message: false
library(lwpls)
library(ggplot2)
```

```{r}
#| label: plot-helpers
#| include: false
theme_article <- function(base_size = 10) {
  theme_bw(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92", linewidth = 0.25),
      panel.border = element_rect(colour = "grey35", linewidth = 0.4),
      strip.background = element_rect(fill = "grey95", colour = "grey35",
                                       linewidth = 0.4),
      axis.ticks = element_line(colour = "grey35", linewidth = 0.3),
      axis.text = element_text(colour = "grey20"),
      legend.background = element_blank(),
      legend.key = element_blank()
    )
}

index_colors <- c("Euclidean" = "#2a78d6", "Covariance-based" = "#eb6834")
weight_colors <- c("white", "#cde2fb", "#86b6ef", "#3987e5", "#1c5cab",
                   "#0d366b")

index_label <- function(similarity) {
  factor(ifelse(similarity == "euclidean", "Euclidean", "Covariance-based"),
         levels = names(index_colors))
}

# Test RMSE versus localization, with the best value of each index and the
# global PLS model for reference.
plot_rmse <- function(curves, global) {
  curves$index <- index_label(curves$similarity)
  best <- do.call(rbind, lapply(split(curves, curves$index), function(d) {
    d[which.min(d$rmse), ]
  }))
  best$label <- sprintf("%.3f", best$rmse)
  ggplot(curves, aes(localization, rmse, colour = index)) +
    geom_hline(yintercept = global, linetype = "dashed", colour = "grey35",
               linewidth = 0.4) +
    annotate("text", x = 1, y = global, vjust = -0.5, size = 3,
             colour = "grey25", label = sprintf("global PLS (%.3f)", global)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.1) +
    geom_point(data = best, size = 3, shape = 21, fill = "white",
               stroke = 0.9) +
    geom_text(data = best, aes(label = label), vjust = 2, size = 3,
              show.legend = FALSE) +
    scale_x_log10() +
    scale_y_continuous(expand = expansion(mult = c(0.14, 0.08))) +
    scale_colour_manual(values = index_colors, name = NULL) +
    labs(x = expression("Localization " * varphi), y = "Test RMSE") +
    theme_article() +
    theme(legend.position = "inside", legend.position.inside = c(0.985, 0.03),
          legend.justification = c(1, 0),
          legend.background = element_rect(fill = "white", colour = NA))
}
```

## One weight function, two distances

For a query sample $x_q$, the training sample $x_i$ receives the weight

$$
\omega_i = \exp\left(-\frac{d_i}{\sigma_d \, \varphi}\right),
$$

where $d_i$ is its distance to the query, $\sigma_d$ is the standard deviation
of the distances of all training samples and $\varphi$ is the `localization`
parameter. Dividing by $\sigma_d$ makes $\varphi$ unitless: a training sample
that is $\varphi$ standard deviations farther from the query than the nearest
one gets $e^{-1} \approx 0.37$ times the weight of the nearest one, whatever
the index (Figure 1). The weights only matter
up to a constant factor, since multiplying all of them by the same number
leaves the weighted PLS model unchanged.

```{r}
#| label: fig-kernel
#| echo: false
#| fig-width: 5.5
#| fig-height: 3
#| fig-cap: "Figure 1. Similarity weight as a function of the distance, for four localization values. Each curve crosses \\(e^{-1}\\) (dotted line) at \\(d_i / \\sigma_d = \\varphi\\) (points)."
#| fig-alt: "Four decreasing exponential curves of the weight against the distance divided by its standard deviation. Smaller localization values decrease faster."
kernel <- expand.grid(u = seq(0, 4, length.out = 400), phi = c(0.25, 0.5, 1, 2))
kernel$weight <- exp(-kernel$u / kernel$phi)
crossings <- data.frame(u = c(0.25, 0.5, 1, 2), phi = c(0.25, 0.5, 1, 2),
                        weight = exp(-1))

ggplot(kernel, aes(u, weight, colour = factor(phi))) +
  geom_hline(yintercept = exp(-1), linetype = "dotted", colour = "grey45",
             linewidth = 0.35) +
  geom_line(linewidth = 0.7) +
  geom_point(data = crossings, size = 1.8) +
  scale_colour_manual(values = weight_colors[3:6], name = expression(varphi)) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0.01, 0.03))) +
  labs(x = expression(d[i] / sigma[d]), y = expression(omega[i])) +
  theme_article() +
  theme(legend.position = "inside", legend.position.inside = c(0.88, 0.68))
```

The two indexes therefore differ only in $d_i$. Both, and the Mahalanobis
distance that we use below for comparison, are quadratic forms of the
difference $\Delta_i = x_i - x_q$ between the standardized predictors:

$$
d_i^2 = \Delta_i^\top M \, \Delta_i,
$$

| Distance | $M$ | Shape of the neighborhood | In lwpls |
|---|---|---|---|
| Euclidean | $I$ | sphere | `similarity = "euclidean"` |
| Mahalanobis | $S^{-1}$ | ellipsoid along the correlations of $X$ | no |
| Covariance-based | $\Gamma \Gamma^\top$ | slab perpendicular to $\Gamma$ | `similarity = "covariance"` |

### Euclidean distance

Write the covariance matrix of the standardized predictors as $S = \sum_k
\lambda_k v_k v_k^\top$, with principal components $v_k$ and variances
$\lambda_k$. The Euclidean distance is

$$
d_i^2 = \lVert \Delta_i \rVert^2 = \sum_k \left(v_k^\top \Delta_i\right)^2 .
$$

Every direction enters the sum with the same coefficient, but for two random
samples the term of component $k$ is $2 \lambda_k$ on average. **The distance
is therefore dominated by the largest sources of variation in $X$**, whether
they are related to the outcome or not. In spectra, these are often baseline
shifts, scattering or interfering compounds. The Euclidean distance does not
use the outcome at all.

The Mahalanobis distance, $M = S^{-1}$, divides each term by $\lambda_k$ so
that every component counts equally. It is often what "covariance-based
distance" means, but it uses the covariance *among the predictors*. It is not
available in lwpls: with more wavelengths than samples $S$ cannot be inverted,
and it inflates the components of smallest variance, which are mostly noise.

### Covariance-based distance

Covariance-based LW-PLS (CbLW-PLS; Hazama and Kano, 2015) uses the covariance
*between the predictors and the outcome*. With centered $X$ and $y$, the unit
vector $w$ that maximizes $\operatorname{cov}(Xw, y) \propto w^\top X^\top y$
is, by the Cauchy--Schwarz inequality $w^\top X^\top y \le \lVert w \rVert \,
\lVert X^\top y \rVert$,

$$
\Gamma = \frac{X^\top y}{\lVert X^\top y \rVert}.
$$

$\Gamma$ is the first weight vector of a PLS model and $t = X\Gamma$ is its
first score. The distance is measured along this direction only:

$$
d_i = \left| \Gamma^\top (x_i - x_q) \right| = |t_i - t_q| .
$$

A one-component PLS model predicts $\hat y = \bar y + b \, t$, so $|\hat y_i -
\hat y_q| = |b| \, d_i$. The factor $|b|$ cancels in $d_i / \sigma_d$: **the
covariance-based weights are those obtained from the distance between the
predictions of a one-component global PLS model.** The neighbors of the query
are the training samples that this simple model predicts to have a similar
outcome.

Since $M = \Gamma\Gamma^\top$ has rank one, all the directions perpendicular
to $\Gamma$ are ignored: samples far from the query in $X$ are full neighbors
if they have the same score $t$. With several outcomes (or classes, coded as
indicators), $\Gamma = X^\top Y / \lVert X^\top Y \rVert_F$ has one column per
outcome and $d_i = \lVert \Gamma^\top (x_i - x_q) \rVert$. Robust models
(`robust = TRUE`) compute $\Gamma$ with the initial weights of partial robust
M-regression, so that outliers do not distort it.

### The three neighborhoods side by side

Two correlated predictors and a nonlinear outcome are enough to see the
shapes:

```{r}
#| label: geometry-data
set.seed(2026)
n <- 250
x1 <- rnorm(n)
x2 <- 0.75 * x1 + sqrt(1 - 0.75^2) * rnorm(n)
y <- sin(1.5 * x1) + 0.3 * x2 + rnorm(n, sd = 0.1)
x <- scale(cbind(x1, x2))

gamma <- crossprod(x, y - mean(y))
gamma <- drop(gamma / sqrt(sum(gamma^2)))
gamma

query <- c(0.6, 0.2)
distance <- list(
  "Euclidean" = function(z) sqrt(rowSums(sweep(z, 2, query)^2)),
  "Mahalanobis" = function(z) sqrt(mahalanobis(z, query, cov(x))),
  "Covariance-based" = function(z) abs(drop(sweep(z, 2, query) %*% gamma))
)
```

```{r}
#| label: fig-geometry
#| echo: false
#| fig-width: 7.2
#| fig-height: 3.75
#| fig-cap: "Figure 2. Weights of the three distances around the same query (white point) with \\(\\varphi = 0.5\\), relative to the largest one, with contour lines at 0.1 and 0.5. The arrow is \\(\\Gamma\\), and \\(n_\\text{eff} = (\\sum \\omega_i)^2 / \\sum \\omega_i^2\\) is the effective number of neighbors."
#| fig-alt: "Three maps of the weights over the two predictors. Euclidean weights form circles around the query, Mahalanobis weights form ellipses along the correlation of the predictors, and covariance-based weights form a band perpendicular to the arrow Gamma."
phi <- 0.5
grid_values <- seq(-3, 3, length.out = 241)
grid <- as.matrix(expand.grid(x1 = grid_values, x2 = grid_values))
panels <- names(distance)

fields <- do.call(rbind, lapply(panels, function(nm) {
  d_train <- distance[[nm]](x)
  d_grid <- distance[[nm]](grid)
  data.frame(grid, index = nm,
             weight = pmin(1, exp(-(d_grid - min(d_train)) / (sd(d_train) * phi))))
}))
# Facet labels with the effective number of neighbors of each distance.
facet_labels <- vapply(panels, function(nm) {
  d_train <- distance[[nm]](x)
  w <- exp(-(d_train - min(d_train)) / (sd(d_train) * phi))
  sprintf('"%s" ~~ (italic(n)[eff] == %.0f)', nm, sum(w)^2 / sum(w^2))
}, character(1))
fields$index <- factor(facet_labels[fields$index], levels = facet_labels)
arrow_df <- data.frame(index = factor(facet_labels[["Covariance-based"]],
                                      levels = facet_labels),
                       x = query[1], y = query[2],
                       xend = query[1] + 1.5 * gamma[1],
                       yend = query[2] + 1.5 * gamma[2])
training <- data.frame(x1 = x[, 1], x2 = x[, 2])

ggplot(fields, aes(x1, x2)) +
  geom_raster(aes(fill = weight), interpolate = TRUE) +
  geom_contour(aes(z = weight), breaks = c(0.1, 0.5), colour = "white",
               linewidth = 0.3) +
  geom_point(data = training, shape = 21, size = 0.9, fill = "grey15",
             colour = "white", stroke = 0.15) +
  geom_segment(data = arrow_df, aes(x = x, y = y, xend = xend, yend = yend),
               linewidth = 0.6, colour = "black",
               arrow = grid::arrow(length = grid::unit(1.8, "mm"), type = "closed")) +
  geom_text(data = arrow_df, aes(x = xend, y = yend), label = "Gamma",
            parse = TRUE, hjust = -0.3, vjust = -0.1, size = 3.6) +
  annotate("point", x = query[1], y = query[2], shape = 21, size = 2.6,
           fill = "white", colour = "black", stroke = 0.7) +
  facet_wrap(~index, nrow = 1, labeller = label_parsed) +
  scale_fill_gradientn(colours = weight_colors, limits = c(0, 1),
                       name = expression(omega[i] / max(omega))) +
  scale_x_continuous(breaks = c(-2, 0, 2)) +
  scale_y_continuous(breaks = c(-2, 0, 2)) +
  coord_equal(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  guides(fill = guide_colourbar(title.position = "top", title.hjust = 0.5)) +
  labs(x = expression(x[1] ~ "(standardized)"), y = expression(x[2] ~ "(standardized)")) +
  theme_article() +
  theme(legend.position = "bottom", legend.key.width = grid::unit(1.6, "cm"),
        legend.key.height = grid::unit(0.25, "cm"), panel.grid = element_blank(),
        panel.spacing = grid::unit(0.8, "lines"),
        legend.margin = margin(0, 0, 0, 0))
```

The Euclidean neighborhood is a circle. The Mahalanobis neighborhood
stretches along the main axis of the cloud, the direction of largest variance
of $X$. The covariance-based neighborhood is a band across that same axis: here
$\Gamma$ points along it, because that is where the predictors covary with
$y$, and differences along $\Gamma$ are exactly what the covariance-based
distance measures.

## When the covariance-based index helps

The simulated spectra below contain three sources of variation: an analyte
band whose absorbance saturates with the concentration (the outcome), so that
the relationship is nonlinear; a large interfering band; and baseline offsets
and slopes. Only the analyte is related to the outcome, while most of the
variance of $X$ comes from the interferent and the baseline.

```{r}
#| label: spectra-data
wavelength <- seq(1100, 1700, by = 5)
band <- function(center, width) exp(-0.5 * ((wavelength - center) / width)^2)

simulate_spectra <- function(n) {
  concentration <- runif(n, 0.05, 1)
  absorbance <- 1 - exp(-2.5 * concentration)
  interferent <- runif(n, 0, 1.5)
  offset <- rnorm(n, sd = 0.15)
  slope <- rnorm(n, sd = 0.1)
  x <- outer(0.8 * absorbance, band(1460, 22)) +
    outer(interferent, band(1330, 70)) +
    outer(offset, rep(1, length(wavelength))) +
    outer(slope, (wavelength - 1400) / 300) +
    matrix(rnorm(n * length(wavelength), sd = 0.004), n)
  colnames(x) <- paste0("nm", wavelength)
  list(x = x, y = concentration)
}

set.seed(7)
train <- simulate_spectra(300)
test <- simulate_spectra(300)
```

We compute $\Gamma$ on the standardized training spectra, as lwpls does, and
measure how well the score $t = X\Gamma$ tracks the outcome:

```{r}
#| label: spectra-gamma
x_train <- scale(train$x)
gamma_spectra <- crossprod(x_train, train$y - mean(train$y))
gamma_spectra <- drop(gamma_spectra / sqrt(sum(gamma_spectra^2)))
cor(drop(x_train %*% gamma_spectra), train$y)
```

```{r}
#| label: fig-spectra
#| echo: false
#| fig-height: 4.6
#| fig-cap: "Figure 3. Top: 40 training spectra, colored by their concentration. Bottom: the covariance direction \\(\\Gamma\\), with the analyte band in gray. \\(\\Gamma\\) peaks on the analyte band; its small, flat loading elsewhere comes from a chance correlation of the outcome with the interferent and the baseline."
#| fig-alt: "Top panel: simulated spectra with a broad interfering band near 1330 nm and a narrower analyte band near 1460 nm. Bottom panel: Gamma, which peaks at the analyte band and is close to zero elsewhere."
shown <- 1:40
spectra_df <- data.frame(
  wavelength = rep(wavelength, each = length(shown)),
  value = as.vector(train$x[shown, ]),
  y = rep(train$y[shown], times = length(wavelength)),
  id = rep(shown, times = length(wavelength)),
  panel = "Absorbance"
)
gamma_df <- data.frame(wavelength = wavelength, value = gamma_spectra,
                       panel = "Gamma[j]")
band_df <- data.frame(wavelength = wavelength,
                      value = band(1460, 22) * max(gamma_spectra),
                      panel = "Gamma[j]")
labels_df <- data.frame(wavelength = c(1330, 1460),
                        value = max(spectra_df$value) * 1.07,
                        label = c("interferent", "analyte"),
                        panel = "Absorbance")
as_panel <- function(df) {
  df$panel <- factor(df$panel, levels = c("Absorbance", "Gamma[j]"))
  df
}
spectra_df <- as_panel(spectra_df)
gamma_df <- as_panel(gamma_df)
band_df <- as_panel(band_df)
labels_df <- as_panel(labels_df)

ggplot(mapping = aes(wavelength, value)) +
  geom_area(data = band_df, fill = "grey88") +
  geom_hline(data = gamma_df[1, ], aes(yintercept = 0), colour = "grey35",
             linewidth = 0.3) +
  geom_line(data = spectra_df, aes(group = id, colour = y), linewidth = 0.35) +
  geom_line(data = gamma_df, colour = index_colors[["Covariance-based"]],
            linewidth = 0.7) +
  geom_text(data = labels_df, aes(label = label), size = 3, colour = "grey25") +
  facet_grid(panel ~ ., scales = "free_y", switch = "y",
             labeller = label_parsed) +
  scale_colour_gradientn(colours = weight_colors[-1], name = "Concentration") +
  labs(x = "Wavelength (nm)", y = NULL) +
  theme_article() +
  theme(strip.placement = "outside", strip.background.y = element_blank(),
        strip.text.y.left = element_text(angle = 90, size = rel(1.1)),
        legend.title = element_text(size = rel(0.9)),
        legend.key.height = grid::unit(0.9, "cm"),
        legend.key.width = grid::unit(0.25, "cm"))
```

We then fit LW-PLS with both indexes over a range of localization values and
compute the test RMSE. A very large localization gives nearly equal weights,
that is, a global PLS model, which we use as a reference:

```{r}
#| label: spectra-rmse
rmse_by_localization <- function(train, test, num_comp, localization) {
  grid <- expand.grid(localization = localization,
                      similarity = c("euclidean", "covariance"),
                      stringsAsFactors = FALSE)
  grid$rmse <- mapply(function(phi, index) {
    fit <- lwpls_fit(train$x, train$y, num_comp = num_comp,
                     localization = phi, similarity = index)
    sqrt(mean((test$y - predict(fit, test$x)$.pred)^2))
  }, grid$localization, grid$similarity)
  grid
}

global_rmse <- function(train, test, num_comp) {
  fit <- lwpls_fit(train$x, train$y, num_comp = num_comp, localization = 1e6)
  sqrt(mean((test$y - predict(fit, test$x)$.pred)^2))
}

localization_grid <- 10^seq(-1.5, 1, length.out = 14)
spectra_rmse <- rmse_by_localization(train, test, num_comp = 4, localization_grid)
spectra_global <- global_rmse(train, test, num_comp = 4)

# Best localization of each index
do.call(rbind, lapply(split(spectra_rmse, spectra_rmse$similarity),
                      function(d) d[which.min(d$rmse), ]))
spectra_global
```

```{r}
#| label: fig-spectra-rmse
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
#| fig-cap: "Figure 4. Test RMSE of LW-PLS with 4 components on the simulated spectra. Open circles mark the best localization of each index, with its RMSE; the dashed line is the global PLS model."
#| fig-alt: "Test RMSE against localization on a log scale. The covariance-based curve stays below the Euclidean curve everywhere, with a minimum about half as large. Both approach the global PLS error for large localization."
plot_rmse(spectra_rmse, spectra_global)
```

The covariance-based index halves the error of the Euclidean index. The
weights of a single query show why. `lwpls_local()` reports the effective
number of neighbors of each prediction, and the weights themselves follow from
the formulas above:

```{r}
#| label: spectra-query
query_id <- which.min(abs(test$y - 0.5))
query_x <- test$x[query_id, , drop = FALSE]
z_query <- (drop(query_x) - attr(x_train, "scaled:center")) /
  attr(x_train, "scaled:scale")

d_euclidean <- sqrt(rowSums(sweep(x_train, 2, z_query)^2))
d_covariance <- abs(drop(sweep(x_train, 2, z_query) %*% gamma_spectra))
relative_weights <- function(d, phi) exp(-(d - min(d)) / (sd(d) * phi))

query_fits <- lapply(c(euclidean = "euclidean", covariance = "covariance"),
                     function(index) {
  lwpls_fit(train$x, train$y, num_comp = 4, localization = 0.2,
            similarity = index)
})
sapply(query_fits, function(fit) lwpls_local(fit, query_x)$reliability$n_eff)
```

```{r}
#| label: fig-spectra-weights
#| echo: false
#| fig-height: 2.9
#| fig-cap: "Figure 5. Weights of the training samples for one test spectrum with a concentration of 0.5 (dashed line), against their own concentration (\\(\\varphi = 0.2\\), weights relative to the largest one)."
#| fig-alt: "Two scatter plots of the weights against the concentration of the training samples. Euclidean weights are high for a few samples spread over all concentrations. Covariance-based weights are zero below a concentration of 0.3 and spread over the samples above."
weights_df <- rbind(
  data.frame(similarity = "euclidean", y = train$y,
             weight = relative_weights(d_euclidean, 0.2)),
  data.frame(similarity = "covariance", y = train$y,
             weight = relative_weights(d_covariance, 0.2))
)
weights_df$index <- index_label(weights_df$similarity)

ggplot(weights_df, aes(y, weight, colour = index)) +
  geom_vline(xintercept = test$y[query_id], linetype = "dashed",
             colour = "grey35", linewidth = 0.4) +
  geom_point(size = 0.9, alpha = 0.75) +
  facet_wrap(~index, nrow = 1) +
  scale_colour_manual(values = index_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(x = "Concentration of the training sample",
       y = expression(omega[i] / max(omega))) +
  theme_article()
```

The Euclidean neighbors are close to the query in interferent and baseline,
not in concentration, so they are spread over the whole concentration range.
The covariance-based index gives almost no weight to the samples below 0.3. Its neighborhood is
still broad and asymmetric, for two reasons: saturation compresses the
differences between high concentrations along $\Gamma$, and $\Gamma$ keeps a
small loading on the interferent and the baseline (Figure 3).

## When the covariance-based index fails

$\Gamma$ only captures the *linear* covariance between the predictors and the
outcome. When the outcome is a symmetric function of a predictor, that
covariance vanishes: with $x_1$ standard normal and $y = x_1^2$,
$\operatorname{cov}(x_1, y) = E[x_1^3] = 0$. $\Gamma$ is then set by chance
correlations in the training set.

```{r}
#| label: symmetric-data
simulate_symmetric <- function(n) {
  x <- matrix(rnorm(n * 5), n, dimnames = list(NULL, paste0("x", 1:5)))
  list(x = x, y = x[, 1]^2 + rnorm(n, sd = 0.2))
}

set.seed(11)
train_sym <- simulate_symmetric(300)
test_sym <- simulate_symmetric(300)

x_sym <- scale(train_sym$x)
gamma_sym <- crossprod(x_sym, train_sym$y - mean(train_sym$y))
gamma_sym <- drop(gamma_sym / sqrt(sum(gamma_sym^2)))
round(gamma_sym, 2)
cor(drop(x_sym %*% gamma_sym), train_sym$y)
```

```{r}
#| label: fig-symmetric
#| echo: false
#| fig-height: 2.9
#| fig-cap: "Figure 6. The outcome against \\(x_1\\), the predictor it depends on (left), and against the score \\(t = X\\Gamma\\) that the covariance-based distance uses (right)."
#| fig-alt: "Left: a clear parabola of the outcome against x1. Right: the outcome against the score t shows no clear pattern."
symmetric_df <- rbind(
  data.frame(panel = "x[1]", coordinate = x_sym[, 1], y = train_sym$y),
  data.frame(panel = "t == X * Gamma", coordinate = drop(x_sym %*% gamma_sym),
             y = train_sym$y)
)
symmetric_df$panel <- factor(symmetric_df$panel,
                             levels = c("x[1]", "t == X * Gamma"))

ggplot(symmetric_df, aes(coordinate, y)) +
  geom_point(size = 0.8, alpha = 0.6, colour = "grey20") +
  facet_wrap(~panel, nrow = 1, labeller = label_parsed) +
  labs(x = "Standardized coordinate", y = "y") +
  theme_article()
```

```{r}
#| label: symmetric-rmse
symmetric_rmse <- rmse_by_localization(train_sym, test_sym, num_comp = 2,
                                       localization_grid)
symmetric_global <- global_rmse(train_sym, test_sym, num_comp = 2)
do.call(rbind, lapply(split(symmetric_rmse, symmetric_rmse$similarity),
                      function(d) d[which.min(d$rmse), ]))
symmetric_global
```

```{r}
#| label: fig-symmetric-rmse
#| echo: false
#| fig-width: 6
#| fig-height: 3.4
#| fig-cap: "Figure 7. Test RMSE of LW-PLS with 2 components for \\(y = x_1^2\\). Open circles mark the best localization of each index, with its RMSE; the dashed line is the global PLS model."
#| fig-alt: "Test RMSE against localization on a log scale. The Euclidean curve reaches a much lower minimum than the covariance-based curve, which stays close to the global PLS error."
plot_rmse(symmetric_rmse, symmetric_global)
```

The Euclidean neighborhoods still follow $x_1$, while the covariance-based
ones mostly follow noise: the covariance-based error is more than twice the
Euclidean one, and only a little below that of the global model. The same happens
whenever the local behavior depends on a direction that the global
covariance $X^\top y$ does not capture, for example regimes defined by other
constituents of the samples (such as matrix effects in emission
spectroscopy).

## Choosing in practice

* **Look at $\Gamma$ first.** Computing it takes one line (see above). If it
  resembles the spectral signature of the property to predict and the score
  $t = X\Gamma$ is well correlated with the outcome, the covariance-based
  index is likely to help. If $\Gamma$ looks like noise, or the correlation is
  weak, prefer the Euclidean index.
* **The covariance-based index suits** data in which most of the variance of
  $X$ is unrelated to the outcome (interferents, baseline, scattering,
  temperature) and the outcome changes monotonically with the relevant
  signal, as in many NIR, Raman or LIBS calibrations.
* **The Euclidean index suits** nonmonotonic relationships, local behavior
  driven by directions other than $X^\top y$, and data whose irrelevant
  variation has already been removed by preprocessing (SNV, derivatives),
  since the Euclidean distance then reflects the relevant variation. It also
  does not depend on the outcome, so it cannot be misled by chance
  correlations.
* **Tune both.** The best localization differs between the indexes (Figures
  4 and 7), so `similarity` should be tuned together with `localization` and
  `num_comp`, as shown in `vignette("lwpls")`.

## References

* Kim, S., Kano, M., Nakagawa, H. and Hasebe, S. (2011). Estimation of active
  pharmaceutical ingredients content using locally weighted partial least
  squares and statistical wavelength selection. *International Journal of
  Pharmaceutics*, 421(2), 269--274.
* Hazama, K. and Kano, M. (2015). Covariance-based locally weighted partial
  least squares for high-performance adaptive modeling. *Chemometrics and
  Intelligent Laboratory Systems*, 146, 55--62.
